College Scorecard Analysis by Justin "Roy" Garrard
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(ggthemes)
library(GGally)
library(plyr)
library(dplyr)
library(Rmisc)
library(scales)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
ed_data <- read.csv("MERGED2014_15_PP.csv")

# Strip the Data Down
# (Location, Admissions, Entry Tests, )
ed_data <- subset(ed_data, ed_data$HIGHDEG >= 4)
ed_data <- subset(ed_data, select = c(UNITID, INSTNM, STABBR, CONTROL, REGION, 
                                      LATITUDE, LONGITUDE, ADM_RATE_ALL, 
                                      TUITIONFEE_IN, TUITIONFEE_OUT, UGDS,
                                      RET_FT4, GRAD_DEBT_MDN, WDRAW_DEBT_MDN,
                                      C150_4))

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Clean_the_Data}
# Clear out a few weird points with a "zero" region
ed_data <- subset(ed_data, ed_data$REGION != 0)

# Remove the University of Phoenix from the data pool, but keep it in a variable
# It's a fascinating outlier that would be worth studying on its own
phoenix <- subset(ed_data, ed_data$UNITID == 484613)
ed_data <- subset(ed_data, ed_data$UNITID != 484613)

# Ensure numeric variables are treated as such
ed_data$TUITIONFEE_IN <- as.numeric(as.character(ed_data$TUITIONFEE_IN))
ed_data$TUITIONFEE_OUT <- as.numeric(as.character(ed_data$TUITIONFEE_OUT))
ed_data$ADM_RATE_ALL <- as.numeric(as.character(ed_data$ADM_RATE_ALL))
ed_data$GRAD_DEBT_MDN <- as.numeric(as.character(ed_data$GRAD_DEBT_MDN))
ed_data$WDRAW_DEBT_MDN <- as.numeric(as.character(ed_data$WDRAW_DEBT_MDN))
ed_data$LONGITUDE <- as.numeric(as.character(ed_data$LONGITUDE))
ed_data$LATITUDE <- as.numeric(as.character(ed_data$LATITUDE))
ed_data$UGDS <- as.numeric(as.character(ed_data$UGDS))
ed_data$RET_FT4 <- as.numeric(as.character(ed_data$RET_FT4))
ed_data$C150_4 <- as.numeric(as.character(ed_data$C150_4))

# Establish certain variables as discrete categories
ed_data$REGION <- factor(ed_data$REGION, ordered=TRUE)
ed_data$CONTROL <- factor(ed_data$CONTROL, ordered=TRUE)

# Change out CONTROL values for understandable strings
ed_data$CONTROL <- revalue(ed_data$CONTROL, c("1"="Public", 
                                              "2"="Private Non-Profit", 
                                              "3"="Private For-Profit"))

# Do some renaming for readability
names(ed_data)[names(ed_data) == 'INSTNM'] <- 'NAME'
names(ed_data)[names(ed_data) == 'STABBR'] <- 'STATE'
names(ed_data)[names(ed_data) == 'CONTROL'] <- 'FUNDING_TYPE'
names(ed_data)[names(ed_data) == 'UGDS'] <- 'UNDERGRAD_ENROLL'
names(ed_data)[names(ed_data) == 'RET_FT4'] <- 'RETENTION'
names(ed_data)[names(ed_data) == 'C150_4'] <- 'COMPLETION_FIVE_YRS'

```

The data explored in this report comes from College Scorecard.
College Scorecard is a product of the U.S. Department of Education
and contains college statistics from 1996 to 2014, though this particular
analysis will only be looking at four-year universities from the 2014 data.

# Univariate Plots Section


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Summary}
# Summary of variables
summary(subset(ed_data, select=c(-UNITID, -NAME)))

```
There are roughly 2000 points of data, each of which was constrained to 
fifteen variables (name and ID number were excluded from the summary).


```{r echo=FALSE, message=FALSE, warning=FALSE, By_Location}
# University by Coordiantes
ggplot(aes(x=LONGITUDE, LATITUDE),
       data=subset(ed_data, ed_data$LONGITUDE < 0 &
                     ed_data$LONGITUDE > -130 &
                     ed_data$LATITUDE > 25)) +
       geom_jitter(alpha=0.25) +
       ggtitle('Mainland U.S. Universities') 


```

```{r echo=FALSE, message=FALSE, warning=FALSE, By_Location_2}
west_count <- subset(ed_data, ed_data$LONGITUDE < -100 &
                       ed_data$LONGITUDE > -130)
east_count <- subset(ed_data, ed_data$LONGITUDE >= -100 &
                       ed_data$LONGITUDE < -65)
non_mainland <- subset(ed_data, ed_data$REGION == 9)
```

The dataset shows that there are approximately four times as many universities 
east of the -100 longitude line than there are west of the line. One interesting
note is that the heat map of university placement closely matches NASA light
pollution images, implying a correlation of universities to urban areas.




```{r echo=FALSE, message=FALSE, warning=FALSE, By_Region}
# University by Region
ggplot(aes(x=REGION), data=ed_data) + geom_histogram(stat="count")

```

The dataset helpfully breaks up location information into nine regions. Excluding
the ninth region (which covers U.S. Territories like Puerto Rico), the fewest
universities are found in region seven. A more visual representation of this
data can be found in the Bivariate Plots section.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, State}
# University by Coordiantes
ggplot(aes(x=STATE),
       data=ed_data) +
       geom_histogram(stat='count') +
       coord_flip() +
       scale_x_discrete(limits=rev(levels(ed_data$STATE))) +
       ggtitle('Universities by State or Territory') 

```

California is the leader in regards to universities per state. New York follows
a reasonable distance behind. Puerto Rico manages to surpass several states, 
while many of the other territories lack even a single university.


```{r echo=FALSE, message=FALSE, warning=FALSE, By_Tuition}
# University by Tuition
in_state <- ggplot(aes(x=TUITIONFEE_IN),
                   data=subset(ed_data, !is.na(ed_data$TUITIONFEE_IN))) +
  geom_histogram(binwidth = 1000) +
  xlab('In-State Tuition in Dollars')

out_state <- ggplot(aes(x=TUITIONFEE_OUT), 
                    data=subset(ed_data, !is.na(ed_data$TUITIONFEE_OUT))) +
  geom_histogram(binwidth = 1000) +
  xlab('Out-Of-State Tuition in Dollars')

multiplot(in_state, out_state)

```
```{r echo=FALSE, message=FALSE, warning=FALSE, By_Tuition_2}
print('In-State Tuition Summary')
summary(ed_data$TUITIONFEE_IN)
print('Out-of-State Tuition Summary')
summary(ed_data$TUITIONFEE_OUT)
```

The median In-State Tuition is slightly above 15,000. The median Out-of-State
Tuition is close to 21,000.


```{r echo=FALSE, message=FALSE, warning=FALSE, Debt}
# University by Debt
grad_debt <- ggplot(aes(x=GRAD_DEBT_MDN),
                    data=subset(ed_data, !is.na(ed_data$GRAD_DEBT_MDN))) +
  geom_histogram(binwidth = 1000) +
  coord_cartesian(xlim=c(0,50000)) +
  xlab('Reported Median Debt for Graduates')

wdraw_debt <- ggplot(aes(x=WDRAW_DEBT_MDN),
                     data=subset(ed_data, !is.na(ed_data$WDRAW_DEBT_MDN))) +
  geom_histogram(binwidth = 1000) +
  coord_cartesian(xlim=c(0,50000)) +
  xlab('Reported Median Debt for Dropouts')

multiplot(grad_debt, wdraw_debt)
```
```{r echo=FALSE, message=FALSE, warning=FALSE, Debt_2}
print('Graduate Debt Summary')
summary(ed_data$GRAD_DEBT_MDN)
print('Dropout Debt Summary')
summary(ed_data$WDRAW_DEBT_MDN)
```

Predictably, graduating students have significantly higher median debt levels
than those who withdraw from university. Also of note is how much more tightly 
clustered and consistent dropout debt is versus graduate debt. 


```{r echo=FALSE, message=FALSE, warning=FALSE, Undergrad_Enrollment}
# University Undergraduate Enrollments
enroll <- ggplot(aes(x=UNDERGRAD_ENROLL), 
                 data=subset(ed_data,
                             !is.na(ed_data$UNDERGRAD_ENROLL) &
                               ed_data$UNDERGRAD_ENROLL > 0)) +
  geom_histogram(binwidth=1000) +
  coord_cartesian(xlim = c(0,50000)) +
  xlab('Reported Enrollment')

enroll_log <- ggplot(aes(x=UNDERGRAD_ENROLL),
                     data=subset(ed_data,
                                 !is.na(ed_data$UNDERGRAD_ENROLL) &
                                   ed_data$UNDERGRAD_ENROLL > 0)) +
  geom_histogram(binwidth=500) +
  coord_cartesian(xlim = c(0,10000)) +
  xlab('Reported Enrollment (0-10000)')

multiplot(enroll, enroll_log)
```
```{r echo=FALSE, message=FALSE, warning=FALSE, Undergrad_Enrollment_2}
print('Undergraduate Enrollment Summary')
summary(ed_data$UNDERGRAD_ENROLL)
max_enroll <- subset(ed_data, ed_data$UNDERGRAD_ENROLL > 10000)
```

Enrollment varies wildly, which is to expected given the numerous sizes of
universities. One notable outlier in the data is the University of Phoenix, 
which is the only university to report a six figure enrollment (151,600). 
This data point has been excluded for most of the analysis in this report.

```{r echo=FALSE, message=FALSE, warning=FALSE, Completion}
# University by Debt
ggplot(aes(x=COMPLETION_FIVE_YRS), 
       data=subset(ed_data,
                   !is.na(ed_data$COMPLETION_FIVE_YRS))) +
  geom_histogram() +
  xlab('Students Who Graduated Within Five Years')
```


```{r echo=FALSE, message=FALSE, warning=FALSE,Completion_3}
print('Undergraduate Five-Year Completion Summary')
summary(ed_data$COMPLETION_FIVE_YRS)
```

Graduation rates follow a surprisingly normal distribution. There are a 
worrisome number of universities (~50) whose completion rates fall below 0.1. 
These values are distinct from NA, so it would seem that they were deliberatly 
reported as such. Of note is their funding types (rarely public) and admission
rates (generally an NA value).

```{r echo=FALSE, message=FALSE, warning=FALSE, Admission_Rate}
# University by Admissions
ggplot(aes(x=ADM_RATE_ALL),
       data=subset(ed_data, !is.na(ed_data$ADM_RATE_ALL))) +
  geom_histogram() +
  xlab('Admission Rate')
```
```{r echo=FALSE, message=FALSE, warning=FALSE, Admission_2}
print('Admission Rates Summary')
summary(ed_data$ADM_RATE_ALL)
```

Admission rates trend towards accepting more often than rejecting, but I imagine
that this varies by other conditions (tution, university funding type, etc.)

```{r echo=FALSE, message=FALSE, warning=FALSE, Retention}
# University by Retention
ggplot(aes(x=RETENTION),
       data=subset(ed_data, !is.na(ed_data$RETENTION))) +
  geom_histogram() +
  xlab('Retention Rate')
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Retention_2}
print('Retention Rates Summary')
summary(ed_data$RETENTION)
no_retention <- subset(ed_data, ed_data$RETENTION < 0.1)
no_ret_or_com <- subset(ed_data, ed_data$RETENTION < 0.1 &
                          ed_data$COMPLETION_FIVE_YRS < 0.1)
```

There is a concerning number of 0.0 and 0.01 retention rates. As before though,
the data draws a distinction between 0.0 and NA, so these were likely reported 
as such. Again, these points are similar in that they're not public universities
and rarely have a listed admission rate.


```{r echo=FALSE, message=FALSE, warning=FALSE, Type}
# University by Type
ggplot(aes(x=FUNDING_TYPE), 
       data=subset(ed_data, !is.na(ed_data$FUNDING_TYPE))) +
  geom_histogram(stat="count") +
  xlab('Universities by Type')

```

Private Non-Profit schools make up a decided majority of the data points.


# Univariate Analysis

### What is the structure of your dataset?

There are roughly 2000 points of data, each of which was constrained to fifteen 
variables. The variables themselves can be grouped into four categories:

* Location (latitude, longitude, state, region)
* Finance (tuitionfee_in, tuitionfee_out, grad_debt_mdn, wdraw_debt_mdn)
* Admission (adm_rate_all, undergrad_enrollment, retention, completion_five_yrs)
* Identification (unitid, name, funding type)

### What is/are the main feature(s) of interest in your dataset?

The primary feature of interest is the completion rate, which offers a 
quantitative view of a university's effectiveness. A low completion rate might 
imply exclusivity, but it also translates into students with debt and nothing to
show for it.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I anticipate that the funding type (public, private) and size (admission rate)
will be strong indicators. Other qualities, such as location and retention, may 
also shed some light on the situation.

### Did you create any new variables from existing variables in the dataset?

No, though it may be handy to have some form of "success quotient" relating
factors such as completion rate and low tuition.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

* University enrollment was adjusted for easier viewing, as there is a
significant spread in the amounts of students admitted.

* Five year completion rates had an unusually normal distribution.

* The ninth region (U.S. territories) and Alaska were excluded from the
longitude/latitude scatter plot. This was done for easier viewing.

* The University of Phoenix was removed from the dataset, as its enrollment was
more than 70 times the median. It's many sub-schools were kept.


# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=15, fig.width=15, Bivariate_Plots}
ggcorr(subset(subset(ed_data,
                       select=c(-STATE, -UNITID, -NAME)),
                ed_data$REGION != 9), label=TRUE, type='lower')

# It's difficult to fit the label names in the plot.
# Seems to be a known issue though.
```

Other than in-state and out-of-state tuition, there are no obvious correlations.
There are a few promising leads though, including tuition/completion rate and 
retention/completion rate.


```{r echo=FALSE, By_Location_and_Region_2}
# University by Coordiantes
ggplot(aes(x=LONGITUDE, LATITUDE),
       data=subset(ed_data,
                   ed_data$LONGITUDE < 0 &
                     ed_data$LONGITUDE > -130 &
                     ed_data$LATITUDE > 25)) +
       geom_jitter(aes(colour=REGION), alpha=0.25) +
       ggtitle('Mainland U.S. Universities') 

```

```{r echo=FALSE, By_Region_2}
# University by Coordiantes
ggplot(aes(x=LONGITUDE, LATITUDE),
       data=subset(ed_data, ed_data$LONGITUDE < 0 & ed_data$LONGITUDE > -130 &
                     ed_data$LATITUDE > 25)) +
       geom_jitter(aes(colour=REGION), alpha=0.25) +
       facet_wrap(~subset(ed_data$REGION, ed_data$LONGITUDE < 0 &
                            ed_data$LONGITUDE > -130 &
                            ed_data$LATITUDE > 25), ncol=3) +
       ggtitle('Mainland U.S. Universities') 

```

Adding color to the geographic maps helps to better demonstrate the shape of the
regions.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=8, Completion_and_Region}
# 
comp_region <- ggplot(aes(x=REGION, y=COMPLETION_FIVE_YRS), data=ed_data) +
       geom_boxplot() +
       theme(axis.title.x=element_blank()) +
       ggtitle('Completion Rate by Region') 

comp_region_2 <- ggplot(aes(x=REGION, y=COMPLETION_FIVE_YRS), data=ed_data) +
       geom_boxplot() +
       coord_cartesian(ylim=c(0.25, 0.75))

multiplot(comp_region, comp_region_2)

```


There's considerable variation between regions with regards to completion rates.
The Northeast regions (1, 2, 3) and West Coast region (8) have a median above
0.5. The Northwest (7), Central (4), and South (5, 6) regions fall below 0.5.
U.S. Territories (9) are particularly afflicted, with a median completion rate
near 0.3.

```{r echo=FALSE, message=FALSE, warning=FALSE, Completion_and_Type}
# 
ggplot(aes(x=FUNDING_TYPE, y=COMPLETION_FIVE_YRS), data=ed_data) +
       geom_boxplot() +
       ggtitle('Completion Rate by University Type') 

```

The funding type of school is shown to be another notable indicator of
completion rate. Private For-Profit universities fall far below Public 
universities in completion rate. Private Non-Profit universities have a 
noticable, if not exceptional, advantage over Public schools.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 6, Completion_and_Tuition}
# 
comp_in <- ggplot(aes(x=TUITIONFEE_IN, y=COMPLETION_FIVE_YRS),
                  data=ed_data) +
       geom_jitter(alpha=0.5) +
       stat_smooth(method = 'lm', col = 'green') +
       ggtitle('Completion Rate by Tuition') 

comp_out <- ggplot(aes(x=TUITIONFEE_OUT, y=COMPLETION_FIVE_YRS),
                   data=ed_data) +
       geom_jitter(alpha=0.5) +
       stat_smooth(method = 'lm', col = 'green')

multiplot(comp_in, comp_out)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Completion_and_Tuition_c}
# 
print('Correlation: In-State Tuition and Completion Rate')
cor_comp_tuit <- subset(ed_data,
                        !is.na(ed_data$TUITIONFEE_OUT) &
                        !is.na(ed_data$COMPLETION_FIVE_YRS))

cor(cor_comp_tuit$TUITIONFEE_IN, cor_comp_tuit$COMPLETION_FIVE_YRS)
print('Correlation: Out-of-State Tuition and Completion Rate')
cor(cor_comp_tuit$TUITIONFEE_OUT, cor_comp_tuit$COMPLETION_FIVE_YRS)
```


Out-of-state tuition appears to be a decent indicator for completion rate. 
In-state tuition follows a similar pattern, but clustering (likely from tuition
subsidies) ruins the trend.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 6, Completion_and_Tuition_2}
# 
ggplot(aes(x=TUITIONFEE_OUT, y=COMPLETION_FIVE_YRS,
           group=floor(TUITIONFEE_OUT/5000)), data=ed_data) +
       geom_boxplot()

```

Binning out-of-state tuition into a box-plot makes the trend a little easier to 
follow.

```{r echo=FALSE, message=FALSE, warning=FALSE, Completion_and_Retention}
# 
ggplot(aes(x=RETENTION, y=COMPLETION_FIVE_YRS),
       data=subset(ed_data, ed_data$RETENTION > 0)) +
       geom_jitter(alpha=0.5) +
       stat_smooth(method = 'lm', col = 'green') +
       ggtitle('Completion Rate by Retention') 

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Completion_and_Retention_c}
# 
print('Correlation: Retention and Completion Rate')
cor_comp_ret <- subset(ed_data,
                        !is.na(ed_data$RETENTION) &
                        !is.na(ed_data$COMPLETION_FIVE_YRS))

cor(cor_comp_ret$RETENTION, cor_comp_ret$COMPLETION_FIVE_YRS)

```

Retention is related to completion rates. This makes sense, since
students not retained cannot graduate (though students who are retained may take
longer than five years to graduate).


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, Completion_and_Exclusivity}
# 
comp_adm <- ggplot(aes(x=ADM_RATE_ALL, y=COMPLETION_FIVE_YRS), data=ed_data) +
       geom_point(alpha=0.5) +
       stat_smooth(method = 'lm', col = 'green') +
       ggtitle('Completion Rate by Admission') 

comp_enroll <- ggplot(aes(x=UNDERGRAD_ENROLL, y=COMPLETION_FIVE_YRS),
                      data=ed_data) +
       geom_point(alpha=0.5) +
       coord_cartesian(xlim=c(0,10000)) +
       stat_smooth(method = 'lm', col = 'green') +
       ggtitle('Completion Rate by Undergraduate Enrollment') 

multiplot(comp_adm, comp_enroll)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Completion_and_Adm_c}
# 
print('Correlation: Admission Rate and Completion Rate')
cor_adm_ret <- subset(ed_data,
                        !is.na(ed_data$ADM_RATE_ALL) &
                        !is.na(ed_data$COMPLETION_FIVE_YRS))

cor(cor_adm_ret$ADM_RATE_ALL, cor_adm_ret$COMPLETION_FIVE_YRS)

print('Correlation: Enrollment and Completion Rate')
cor_en_ret <- subset(ed_data,
                        !is.na(ed_data$UNDERGRAD_ENROLL) &
                        !is.na(ed_data$COMPLETION_FIVE_YRS))

cor(cor_en_ret$UNDERGRAD_ENROLL, cor_en_ret$COMPLETION_FIVE_YRS)

```

Neither admission rate or undergraduate enrollment seem to be related to
completion rate. 


```{r echo=FALSE, message=FALSE, warning=FALSE, Completion_and_Debt}
# 
ggplot(aes(x=GRAD_DEBT_MDN, y=COMPLETION_FIVE_YRS), data=ed_data) +
       geom_jitter(alpha=0.5) +
       ggtitle('Completion Rate by Median Graduate Debt') 

```

The median debt of graduating students does not appear to have any discernable
relationship to completion rate.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Out-of-state tuition, student retention, the funding type of university, and the
region of the university all seem to have some relationship to completion rates.
The size of the university and its exclusivity (admission rate) have less 
bearing. Likewise, the median debt of a graduate seem unrelated to completion 
rate.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

In-state tuition and out-of-state tuition are correlated, which is to be 
expected, but the shape of their scatter plots offers some interesting insight
into how tuition rates are set. There's a strong clustering around the \$7,000
mark and a break before the \$20,000 mark. Following \$20,000, in-state tuition 
looks nearly identical to out-of-state tuition. This shape implies that 
universities drawinga distinction between in-state and out-of-state tuition have
rates that are less than \$20,000.

Median graduate debt also exhibits a strange patterning. Hard lines exist
at the \$25,000 and \$27,000 values, utterly independent of completion rate.
This suggests a "standard value" of sorts that universities and financial aid
packages aim for.

### What was the strongest relationship you found?

The funding type of university has a noticable relationship to its completion
rate. For-profit universities have, on average, lower completion rates than
either public or non-profit universities.

# Multivariate Plots Section


```{r echo=FALSE, message=FALSE, warning=FALSE, Type_Completion_Tuition}

ggplot(aes(x=TUITIONFEE_OUT, y=COMPLETION_FIVE_YRS, colour=FUNDING_TYPE),
       data=ed_data) +
       geom_jitter(alpha=0.5)

```

```{r echo=FALSE, fig.height=5, fig.width=10, message=FALSE, warning=FALSE, Type_Completion_Tuition_Region}

ggplot(aes(x=TUITIONFEE_OUT, y=COMPLETION_FIVE_YRS, colour=FUNDING_TYPE),
       data=ed_data) +
       geom_jitter(alpha=0.5) +
       facet_wrap(~REGION)

```

Adding funding type and region to the plot outlines an interesting divide
between universities. For-Profit universities make up the lowest completion
rates and tuitions. Non-Profit universities comprise the majority of 
high-tuition, high-completion rate schools. Public universities generally fall 
in-between.

Each region seems to experience this discrepency in a different way. Region 9
(U.S. Territories) has a surprising number of non-profit universities. Region 8
(West Coast) seems to go against the trend with an odd patterning of public
and non-profit schools. Region 5 (Southeast) has so many universities that it's
hard to see a trend.


```{r echo=FALSE, fig.width=8, message=FALSE, warning=FALSE, Type_Completion_Retention}
# 
ggplot(aes(x=RETENTION, y=COMPLETION_FIVE_YRS, colour=FUNDING_TYPE),
       data=subset(ed_data, ed_data$RETENTION > 0)) +
       geom_jitter(alpha=0.5) +
       facet_wrap(~FUNDING_TYPE) +
       stat_smooth(method = 'lm', col = 'green') +
       ggtitle('Completion Rate by Retention') 

```

Categorizing funding types by color gives this plot a new story to tell. Public
universities manifest the most consistent relationship while non-profits are 
more erratic. For-profit universities, confusingly, show no pattern. That would 
imply that there is no connection between the number of students that graduate 
in five years and the number of students that choose to stay each year.


```{r echo=FALSE, message=FALSE, warning=FALSE, Type_Completion_Debt}
# 
ggplot(aes(x=GRAD_DEBT_MDN, y=COMPLETION_FIVE_YRS, colour=FUNDING_TYPE),
       data=ed_data) +
       geom_jitter(alpha=0.5) +
       ggtitle('Completion Rate by Median Graduate Debt') 

```

```{r echo=FALSE, fig.width=10, message=FALSE, warning=FALSE, Type_Completion_Debt_2}
# 
ggplot(aes(x=GRAD_DEBT_MDN, y=COMPLETION_FIVE_YRS, colour=FUNDING_TYPE),
       data=ed_data) +
       geom_jitter(alpha=0.5) +
       facet_wrap(~FUNDING_TYPE) +
       ggtitle('Completion Rate by Median Graduate Debt') 

```
This plot had previously been an enigma. Looking through the lens of 
funding type, however, shows useful patterns. Public and non-profit universities
have a fairly similar structure with graduate debt. Non-profits in particular
show hard lines at 25,000 and 27,000.

For-profit universities show the highest amount of debt. I find this surprising,
given it was previously established that they have lower tuition rates on
average.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Tying together out-of-state tuition, university funding type, and region proved
a very effective means for understanding completion rates. Independently they
have weak relationships to one another, but when drawn together, they show some
remarkable patterns. 

### Were there any interesting or surprising interactions between features?

That for-profit schools have some of the lowest tuitions but highest graduate
debts was unexpected (and worrisome). I would be curious to discover why this
is the case. Perhaps there are financial aid restrictions in play? Or is it
the demographics?

For-profit schools seem strange on the whole. In graduate debt and retention,
they follow completely different patterns than other universities. Is this
because of differences in student demographic? Or something else entirely?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.


N/A

------

# Final Plots and Summary


### Plot One
```{r echo=FALSE,  Plot_One}
# University by Coordiantes
ggplot(aes(x=LONGITUDE, LATITUDE),
       data=subset(ed_data, ed_data$LONGITUDE < 0 &
                     ed_data$LONGITUDE > -130 &
                     ed_data$LATITUDE > 25)) +
       geom_jitter(aes(colour=REGION), alpha=0.25) +
       ggtitle('Mainland U.S. Universities') 
```

### Description One

This plot simulataneously outlines the relative proportions of universities in
the U.S. while providing context for the "Region" variable that's referenced
throughout. Readers can quickly deduce that the majority of universities exist
in the Eastern U.S. They can also see the vivid outline of the West Coast and
blatant sparsity of the Midwest.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=7, fig.width=12, Plot_Two}
ggplot(aes(x=TUITIONFEE_OUT, y=COMPLETION_FIVE_YRS, colour=FUNDING_TYPE),
       data=ed_data) +
       geom_jitter(alpha=0.5) +
       facet_wrap(~REGION) +
       xlab('Out-of-State Tuition (USD)') +
       ylab('Percentage of Students Graduating Within Five Years') +
       ggtitle('Regional Map of Five Year Completion vs. Tuition') +
       scale_x_continuous(labels = dollar)
```

### Description Two

There's a lot going on in this map, but it demonstrates some crucial
relationships. First, it shows the relationship between university completion
rate and tuition. Some regions demonstrate this more than others, but the 
overall trend can be observed. Second, it illustrates a national pattern where 
public, non-profit, and for-profit schools exist on a spectrum. Dividing the map
by region mitigates the number of outliers viewers have to sift through. Lastly,
the chart juxtoposes the various regions with one another. It's easy to see at a
glance that regions 9 and 7 have fewer universities than 5 or 2.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 6, Plot_Three}
comp_in <- ggplot(aes(x=TUITIONFEE_IN, y=COMPLETION_FIVE_YRS),
                  data=ed_data) +
       geom_jitter(alpha=0.5) +
       stat_smooth(method = 'lm', col = 'green') +
       ggtitle('Completion Rate by Tuition') +
       xlab('In-State Tuition (USD)') +
       ylab('Five Year Completion Rate') +
       scale_x_continuous(labels = dollar)

comp_out <- ggplot(aes(x=TUITIONFEE_OUT, y=COMPLETION_FIVE_YRS),
                   data=ed_data) +
       geom_jitter(alpha=0.5) +
       stat_smooth(method = 'lm', col = 'green') +
       xlab('Out-of-State Tuition (USD)') +
       ylab('Five Year Completion Rate') +
       scale_x_continuous(labels = dollar)

multiplot(comp_in, comp_out)
```

### Description Three

This plot was an unexpected surprise. It shows the overall pattern of completion
rate and tution fees, but also the effect of in-state tuition. There isn't a
simple left-shift like what one might expect; rather, it's as though there's
some gravitational force pulling points towards the \$5,000 mark.

It's also interesting to note that the plot is relatively unaffected past the
\$20,000 mark. Given the similarity in shapes between the two graphs, one
can deduce that few schools beyond that point offer different in-state tuitions.

------

# Reflection

The College Scorecard is a vast repostiory of information across several years.
This analysis covered data for four-year universities in the year 2014. I placed
an emphasis on fifteen variables in particular, which detailed location,
admissions, finances, and identification.

My focus was on investigating the completion rate of various universities.
Tuition rates, university funding types, and region were found to be
influential. The most significant influence came from the university funding
type, as for-profit schools on average have half the completion rate of 
non-profit schools. Similarly, higher tuition rates seem to go hand-in-hand with
higher completion rates.

There was some difficulty in importing, cleaning, and understanding the College
Scorecard data. The data itself is massive; my spreadsheet program could not
open the .csv because there were simply too many columns. Many of the column
names were unintuitive as well (C150_4 comes to mind). 

I was pleased with how questions raised by the Bivariate Plots were answered by
the Multivariate Plots. Looking at the data through the lens of university
funding type showed just how distinct each group is. Future work could certainly
be done with regards to investigating the patterns within each funding type, and
across multiple years.

